flowCore: data structures package for flow cytometry data

### 2009-03-24 inst/doc/HowTo-flowCore.Rnw
edits to the vignette
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@35702 bc3139a8-67e5-0310-9ffc-ced21a209358
reduce the package dependencies and optimize the search path strategy
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@37386 bc3139a8-67e5-0310-9ffc-ced21a209358
update for revised version of the manuscript: logicle reference, flowCore vignette...
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@38141 bc3139a8-67e5-0310-9ffc-ced21a209358

% NOTE -- ONLY EDIT THE .Rnw FILE!!!  The .tex file is
% likely to be overwritten.
%
%\VignetteIndexEntry{Basic Functions for Flow Cytometry Data}
%\VignetteDepends{flowCore}
%\VignetteKeywords{}
%\VignettePackage{flowCore}
\documentclass[11pt]{article}

@@

Though any function can be used as a transform in both the out-of-line
and inline transformation techniques, \Rpackage{flowCore} provides a
number of parameterized transform generators that correspond to the
transforms commonly found in flow cytometry and defined in the
Transformation Markup Language (Transformation-ML). Briefly, the
predefined transforms are:

@@

\section{Filtering}

The most common task in the analysis of flow cytometry data is usual
some form of filtering operation, either to obtain summary statistics
about the number of events that meet a certain criteria or to perform
further analysis on a subset of the data.

\subsection{Standard Filters}

Most filtering operations are a composition of one or more common
filtering operations. Like transformations, \Rpackage{flowCore}
includes a number of built-in common flow cytometry filters.

The simplest of these filters are the geometric filters, which
correspond to those typically found in interactive flow cytometry
software:
\begin{description}
\item[rectangleGate]{Describes a cubic shape in one or more
    dimensions--a rectangle in one dimension is simply an interval
    gate.}
\item[polygonGate]{Describes an arbitrary two dimensional polygonal
    gate.}
\item[polytopeGate]{Describes a region that is the convex hull of the
    given points. This gate can exist in dimensions higher than $2$,
    unlike the \Robject{polygonGate}.}
\item[ellipsoidGate]{Describes an ellipsoidal region in two or more
    dimensions}
\end{description}

@@

We also include some data-driven filters not usually found in flow
cytometry software:

@@

Transformation operations (and also compensations or normalizations)
only create single \Rclass{views}. This is not true for gating
operations. Even the most simple gate creates two populations: those
cells in the gate and the complement. Accordingly, two separate
\Rclass{views} have to be created as well. 
<<addGate>>=
rg <- rectangleGate("FSC-H"=c(200,400), "SSC-H"=c(250, 400),
                    filterId="rectangle")
add(wf, rg, parent="asinh")
wf
@
%
As we see, there are now two new \Rclass{views} under the ``asinh''
node called ``rectangle+'' and ``rectangle-''. Other filter types may
create more than two populations, for instance a \Rclass{quadrantGate}
always results in four sub-populations. There is no restriction on the
number of populations that are allowed to be created by a single
gating operation, however, when gating a \Rclass{flowSet}, the number
of populations have to be the same for each frame in the set.
<<addQuadGate>>=
qg <- quadGate("FL1-H"=2, "FL2-H"=4)
add(wf,qg,parent="rectangle+")
wf
@ 
%
For complex work flows it becomes increasingly hard to convey all
information by printing on the screen. There is a plotting function
for \Rclass{workFlow} objects which plots the underlying tree. The
function depends on the \Rpackage{Rgraphviz} package.
<<plotwf, fig=true>>=
plot(wf)
@

>>>>

% NOTE -- ONLY EDIT THE .Rnw FILE!!!  The .tex file is
% likely to be overwritten.
%
%\VignetteIndexEntry{Basic Functions for Flow Cytometry Data}
%\VignetteDepends{flowCore, flowViz}
%\VignetteKeywords{}
%\VignettePackage{flowCore}
\documentclass[11pt]{article}

@@

Though any function can be used as a transform in both the out-of-line
and inline transformation techniques, \Rpackage{flowCore} provides a
number of parameterized transform generators that correspond to the
transforms commonly found in flow cytometry and defined in the
Transformation Markup Language (Transformation-ML, see \url{http://www.ficcs.org/} and \cite{Spidlen2006} for more details). Briefly, the
predefined transforms are:

@@

\section{Gating}

The most common task in the analysis of flow cytometry data is some
form of filtering operation, also known as gating, either to obtain
summary statistics about the number of events that meet a certain
criteria or to perform further analysis on a subset of the data.  Most
filtering operations are a composition of one or more common filtering
operations. The definition of gates in \Rpackage{flowCore} follows the
Gating Markup Language Candidate Recommendation \cite{Spidlen2008},
thus any \Rpackage{flowCore} gating strategy can be reproduced by any
other software that also adheres to the standard and \textit{vice
  versa}.

\subsection{Standard gates and filters}
Like transformations, \Rpackage{flowCore} includes a number of
built-in common flow cytometry gates. The simplest of these gates are
the geometric gates, which correspond to those typically found in
interactive flow cytometry software:
\begin{description}
\item[rectangleGate]{Describes a cubic shape in one or more
    dimensions--a rectangle in one dimension is simply an interval
    gate.}
\item[polygonGate]{Describes an arbitrary two dimensional polygonal
    gate.}
\item[polytopeGate]{Describes a region that is the convex hull of the
    given points. This gate can exist in dimensions higher than $2$,
    unlike the \Robject{polygonGate}.}
\item[ellipsoidGate]{Describes an ellipsoidal region in two or more
    dimensions}
\end{description}

@@

In addition, we introduce the notion of data-driven gates, or filters,
not usually found in flow cytometry software. In these approaches, the
necessary parameters are computed based on the properties of the
underlying data, for instance by modelling data distribution or by
density estimation :

@@

Transformation operations (and also compensations or normalizations)
only create single \Rclass{views}. This is not true for gating
operations. Even the most simple gate creates two populations: those
cells in the gate and the complement. Accordingly, two separate
\Rclass{views} have to be created as well. 
<<addGate>>=
rg <- rectangleGate("FSC-H"=c(200,400), "SSC-H"=c(250, 400),
                    filterId="rectangle")
add(wf, rg, parent="asinh")
wf
@
%
As we see, there are now two new \Rclass{views} under the ``asinh''
node called ``rectangle+'' and ``rectangle-''. Other filter types may
create more than two populations, for instance a \Rclass{quadrantGate}
always results in four sub-populations. There is no restriction on the
number of populations that are allowed to be created by a single
gating operation, however, when gating a \Rclass{flowSet}, the number
of populations have to be the same for each frame in the set.
<<addQuadGate>>=
qg <- quadGate("FL1-H"=2, "FL2-H"=4)
add(wf,qg,parent="rectangle+")
wf
@ 
%
For complex work flows it becomes increasingly hard to convey all
information by printing on the screen. There is a plotting function
for \Rclass{workFlow} objects which plots the underlying tree. The
function depends on the \Rpackage{Rgraphviz} package.
<<plotwf, eval=FALSE>>=
plot(wf)
@ 
<<plotwfdo, fig=true, echo=FALSE, results=hide>>=
if(suppressWarnings(require(Rgraphviz))){
    plot(wf)
}else{
    plot(1,1, type="n", axes=FALSE, ann=FALSE)
    text(1,1,"Need to install Rgraphviz")
}
@
